Economic Drivers, Immigration Dynamics, and Housing Starts: Their Collective Influence on Housing Price Index

### 2024-04-19 paper/paper.qmd
added remaining scripts, final paper changes, deleted files

thanks: "Code and data are available at: https://github.com/hoodanav/Housing_Market_Analysis/tree/main"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

Focusing particularly on Ontario's housing landscape, where activity has surged in recent years, this paper investigates shifts in housing demand over the past two decades. We evaluate whether current housing development initiatives adequately address the housing crisis. Additionally, we discuss government policies regarding housing capacity and existing targets.

Through this analysis, we aim to grasp the current reality of the housing market and understand how economic conditions, including factors like immigration and housing starts, influence its trajectory. With the housing market facing constraints such as high interest rates, we seek to uncover insights regarding the future of the market.

The remainder of this paper is structured as follows.In the Data@sec-data, we note the sources of the data sets utilized in the paper, the specific techniques we used to process them, and key variables we wish to use. In the Model @sec-data2, we created models to assess the impact of the key variables on our outcome of housing prices. In the Results @sec-data3 we discuss  the results of the model and what key correlations were found from the variables studied in the paper. In the Discussion @sec-data4, we explore the findings of modeling and predictions. In addition, we discuss biases and weaknesses in the data that contributed to our findings, and how we navigated those limitations.

@@

```{r}
# Load necessary libraries
library(dplyr)
library(knitr)
library(kableExtra)

# Display the first 6 rows of the merged data as a table using kable with proper styling
kable(head(merged_data, 6), format = "latex", caption = "Merged Data Table") %>%
  kable_styling(latex_options = "scale_down") %>%
  as.character() %>%
  cat("\n\\centering")  # Add LaTeX centering command

@@

@@

## Data Modifications
Overall, the values were often reported at different times, for consistency we chose to use a yearly metric. We averaged upon quarterly or monthly reported data to gain an averaged yearly value. We then created a merged datadframe joined by the year variable and achieved our cleaned data as seen in @tbl-clean-data.

```{r}
# Load the necessary packages
#| label: fig-immigration-plot
#| fig-cap: Number of immigrants in Canada from 2000 to 2023
#| echo: false
#| warning: false
#| message: false
# Filter out rows with missing values in the 'Number.of.immigrants.in.Canada.from.2000.to.2023' column
merged_data_filtered <- merged_data[!is.na(merged_data$Number.of.immigrants.in.Canada.from.2000.to.2023),]

# Create a line plot for the number of immigrants in Canada over time
ggplot(merged_data_filtered, aes(x = Year, y = `Number.of.immigrants.in.Canada.from.2000.to.2023`)) +
  geom_line(color = "blue") +
  labs(title = "Number of New Immigrants in Canada from 2000-2023",
       x = "Year",
       y = "Number of Immigrants (Thousands)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +  # Center the title
  scale_y_continuous(labels = scales::comma_format(scale = 1e-3), limits = c(0, NA))  # Format y-axis labels

@@

```{r}
#| label: fig-housing-plot
#| fig-cap: Housing Starts Over 2000-2023
#| echo: false
#| warning: false
#| message: false

@@

# Plot housing starts over time with adjusted y-axis scale
plot(housing_starts$Year, housing_starts$Starts, type = "l", 
     xlab = "Year", ylab = "Housing Starts", 
     main = "Housing Starts Over Time",
     ylim = c(0, max_starts_rounded),
     yaxp = c(0, max_starts_rounded/1000, 10),  # Adjusting the number of ticks on the y-axis
     col = "blue")  # Set the line color to blue

@@

```{r}
#| label: fig-hpivs-plot
#| fig-cap: Number of immigrants in Canada from 2000 to 2023
#| echo: false
#| warning: false
#| message: false
# Scatterplot of HPI against each economic indicator
library(ggplot2)

ggplot(data = merged_data, aes(x = Interest_Rate, y = Inflation.Rate....)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Adding a linear trend line
  labs(title = "Scatterplot of HPI vs Interest Rate",
       x = "Interest Rate",
       y = "HPI")

@@

The goal of our modeling strategy is to analyze the relationship between the economic predictors and the housing price index (HPI), and explore how the economic variables contribute to the housing market. After exploring some of the intial data we observed a general trend between different variables, that indicated some level of relationship between them. To see the actual relationships, we implement a multiple linear regression model to take a close look at what variables seem to have an effect on the outcome variable, denoted by the house price index.

### Multiple Linear Regression
We chose to explore a variety of economic and count variables like immigration and housing starts, to see how these impact the housing prices we use multiple linear regression model.

@@

\[ \text{House\_Price\_Index} = \beta_0 + \beta_1 \times \text{Inflation\_Rate} + \beta_2 \times \text{Interest\_Rate} + \beta_3 \times \text{Unemployment\_Rate} + \beta_4 \times \text{Immigration\_numbers} + \beta_5 \times \text{Housing\_Starts} + \epsilon \]

@@

Our multiple linear regression model is summarized in [@tbl-model-sum]. Our multiple linear regression model reveals the strength and association of each indicator to the outcome variable, while holding other variables constant. The estimated coefficients in the model indicate the magnitude and direction of each variable's association. Indicators that had a positive relationship include the inflation_rate where each unit of increase in its value correlated to a 5.398 unit (estimate coefficient) increase in house_price_index.

The variables interest rate and unemployment_rate had a negative association to the house_price_index, as expected.For each unit increase of the interest rate (percent) the house price index saw a 12.45 unit decrease. The unemployment rate had estimate coefficient -2.812 indicating that as the unemployment rate changed by a unit the house price index saw a 2.812 unit decrease. However, here the relationship is not statistically significant as p > 0.05.

@@

The model demonstrates an overall good fit with an R-squared value of 0.8522, indicating that approximately 85.229% of the variance in house price index is explained by the included predictors. The adjusted R-squared, which accounts for the number of predictors in the model, is slightly lower at 0.8088, suggesting the model is fairly robust. The residual standard error is 8.364, suggesting the amount of variability of observed responses around the fitted regression line (plot).

@@

## Interest Rates and HPI
The model suggests interest rate as a significant predictor of HPI

@@

>>>>

thanks: "Code and data are available at: https://github.com/hoodanav/Housing_Market_Analysis/tree/main"
date: today
date-format: long
abstract: "This paper examines Ontario's housing market over 20 years, focusing on economic variables, housing starts, and immigration's impact on house prices. We evaluate the current housing development initiatives in addressing the housing crisis, along with immigration targets as supply and demand variables. The market's current state indicates an already constricted inventory that is largely influenced by interest rate hikes over the last few years."
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
format: pdf 
number-sections: true
bibliography: references.bib
toc: true
---

@@

Focusing particularly on Ontario's housing market, where activity has surged in recent years, this paper investigates shifts in housing demand over the past two decades. We evaluate whether current housing development initiatives adequately address the housing crisis. Additionally, we discuss government policies regarding housing capacity and existing targets.

Through this analysis, we aim to grasp the current reality of the housing market and understand how economic conditions, including factors like immigration and housing starts, influence its trajectory. With the housing market facing constraints such as high interest rates, we found that they did have a significant correlation to changes in house price index, whereas certain factors did not have much association to house price index at all.

The remainder of this paper is structured as follows.In the Data @sec-data, we note the sources of the data sets utilized in the paper, the specific techniques we used to process them, and key variables we wish to use. In the Model @sec-data2, we created models to assess the impact of the key variables on our outcome of housing prices. In the Results @sec-data3 we discuss  the results of the model and what key correlations were found from the variables studied in the paper. In the Discussion @sec-data4, we explore the findings of modeling and predictions. In addition, we discuss biases and weaknesses in the data that contributed to our findings, and how we navigated those limitations.

@@

```{r} 
#| label: tbl-cleaned-data
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: "Clean and merged data"
library(kableExtra)

options(kableExtra.latex.max.width=1)
head(merged_data, 6) |>
  kable(
        col.names = c("Year","Average_House_Only","Interest_Rate","Unemployment Rate (%)","Annual Change.x","Inflation Rate (%)","Annual Change.y","Number of immigrants in Canada from 2000 to 2023","Starts"),
    booktabs = TRUE,
    longtable = FALSE
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"), font_size = 3)

@@

```{r}
#| label: tbl-model-sum
#| fig-cap: "Model Summary"
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(dplyr)
library(knitr)
library(kableExtra)
library(here)

# Load the model file
pricing_model <-
  readRDS(file = here::here("models/pricing_model.rds"))

# # Now you can access model_data
# # For example, if model_data is a data frame in the model file
# 
# 
# # Create a function to format the coefficients table
# format_coefficients_table <- function(model_summary) {
#   # Extract coefficients, standard errors, t-values, and p-values
#   coefficients <- coef(model_summary)
#   std_errors <- summary(model_summary)$coefficients[, "Std. Error"]
#   t_values <- summary(model_summary)$coefficients[, "t value"]
#   p_values <- summary(model_summary)$coefficients[, "Pr(>|t|)"]
#   
#   # Create a data frame
#   coefficients_table <- data.frame(
#     Coefficient = coefficients,
#     `Std. Error` = std_errors,
#     `t value` = t_values,
#     `Pr(>|t|)` = p_values
#   )
#   
#   return(coefficients_table)
# }
# 
# # Call the function with your model summary
# model_summary <- lm(House_Price_Index ~ Inflation_Rate + Interest_Rate + 
#                       Unemployment_Rate + Immigration_numbers + Housing_Starts, 
#                     data = model_data)
# coefficients_table <- format_coefficients_table(model_summary)
# 
# # Print the coefficients table using kable and add formatting
# kable(coefficients_table, format = "markdown", align = "c") %>%
#   kable_styling(latex_options = c("striped", "scale_down")) %>%
#   add_header_above(c(" " = 1, "Coefficients" = 4))

```{r}
#| label: tbl-model
#| tbl-cap: Model Summary Results 
#| echo: false
#| warning: false
#| message: false
modelsummary::modelsummary(
  list(
    "Price model" = pricing_model
  ),
  fmt = 2
)

@@

## Data Modifications
Overall, the values were often reported at different times, for consistency we chose to use a yearly metric. We averaged upon quarterly or monthly reported data to gain an averaged yearly value. We then created a merged datadframe joined by the year variable and achieved our cleaned data as seen in @tbl-cleaned-data.

@@

```{r}
#| label: fig-housing-plot
#| fig-cap: Housing Starts Over 2000-2023 in Ontario
#| echo: false
#| warning: false
#| message: false

@@

# Plot housing starts over time with adjusted y-axis scale
plot(housing_starts$Year, housing_starts$Starts, type = "l", 
     xlab = "Year", ylab = "Housing Starts", 
     main = "",
     ylim = c(0, max_starts_rounded),
     yaxp = c(0, max_starts_rounded/1000, 10),  # Adjusting the number of ticks on the y-axis
     col = "blue")  # Set the line color to blue

@@

@@

The goal of our modeling strategy is to analyze the relationship between the economic predictors and the housing price index (HPI), and explore how the economic variables contribute to the housing market. After exploring some of the initial data we observed a general trend between different variables, that indicated some level of relationship between them. To see the actual relationships, we implement a bayesian generalized linear model to take a close look at what variables seem to have an effect on the outcome variable, denoted by the house price index.

###  Bayesian linear regression
We chose to explore a variety of economic and count variables like immigration and housing starts, to see how these impact the housing prices we use a Bayesian GLM.

@@

$$
\begin{aligned}
\text{House\_Price\_Index} = & \beta_0 + \beta_1 \times \text{Inflation\_Rate} \\
& + \beta_2 \times \text{Interest\_Rate} + \beta_3 \times \text{Unemployment\_Rate} \\
& + \beta_4 \times \text{Immigration\_numbers} + \beta_5 \times \text{Housing\_Starts} + \epsilon
\end{aligned}
$$

@@

Our multiple linear regression model is summarized in [@tbl-model]. Our multiple linear regression model reveals the strength and association of each indicator to the outcome variable, while holding other variables constant. The estimated coefficients in the model indicate the magnitude and direction of each variable's association. Indicators that had a positive relationship include the inflation_rate where each unit of increase in its value correlated to a 3.046 unit (estimate coefficient) increase in house_price_index.

The variables interest rate and unemployment_rate had a negative association to the house_price_index, as expected.For each unit increase of the interest rate (percent) the house price index saw a 11.579 unit decrease. The unemployment rate had estimate coefficient -2.320 indicating that as the unemployment rate changed by a unit the house price index saw a 2.320 unit decrease. However, here the relationship is not statistically significant.

@@

The model demonstrates an overall good fit with an R-squared value of 0.842, indicating that approximately 84.2% of the variance in house price index is explained by the included predictors. The adjusted R-squared, which accounts for the number of predictors in the model, is slightly lower at 0.745, suggesting the model is fairly robust. The log likelihood has a value of -74.00 and indicates that the model provides a relatively good fit, as this refers to how well the model explains the observed data.

@@

## Interest Rates and HPI
The model suggests interest rate as a significant predictor of HPI, or one that has the most fluctuation in a one unit change to our response variable of house price index. This trend was expected and is justified considering economic tendencies. The interest rates in our study recently saw a consistent hike over the last few years. This interest rate is the measure of the mortgage lending rate, a variable directly correlated to housing affordability and prices. As these have gone up, this negative relationship towards the house price index is justified. In terms of the housing market, interest rates show to be beneficial for generally decreasing house price index, and likely other market predictors not evaluated in this study.

@@

The
